Supplementary For\\ MGRFE: multilayer recursive feature elimination based on an embedded genetic algorithm for cancer classification

### 2018-12-01 MGRFE-TCBB-2018-08-16/supplementary.tex
Complete! Record my gratitude.

	We implemented the proposed MGRFE in Python version 3.6.0 environment (\url{https://www.python.org/}) on a common laptop computer with Intel(R) Core(TM) i5-4210U CPU and 8G memory. The Python SciPy package version 0.19.0 \cite{RN441} was involved in the \textit{t}-test process, and minepy package version 1.2.0 \cite{RN440} was used to perform the MIC calculation. Some parameter settings about MGRFE: 1) the evolution iteration number of embedded GA was dynamically set to 1 to 3 (smaller iteration number used for larger chromosome length to save time); 2) the reduced feature number between two GA runs, the RFE step, was dynamically set to 1 to 3 (larger reduction step used for larger chromosome length to save time); and 3) the iterative layer number of MGRFE being 3 with three, two and one GaRFE processes at each layer is usually enough. In the experiments, we limited the size of the final selected gene combination in each dataset to below 10 genes. According to the experiment records, in each of the 19 microarray datasets, the running time of MGRFE is commonly between 500 seconds (8.33 minutes) and 900 seconds (15 minutes). Additionally, it is well worthy to mention that the final chosen gene subset in each dataset might be already found by the first GaRFE process in the first layer of MGRFE, which just costs 2\(\scriptsize{\sim}\)3 minutes. More implementation details and experiment results of MGRFE in the 19 datasets are available at \url{https://github.com/Pengeace/MGRFE-GaRFE}.

	Because Kar \emph{et~al.} also employed an evolutionary-computation method PSO, which is similar to GA, to select minimal informative genes in microarray and provided their program running time records on three datasets SRBCT, ALL\_AML, and MLL \cite{W15}, here, we offer a simple running time comparison between their method and MGRFE. Their PSO-based method cost 2.7956, 2.7906 and 7.1488 hours on SRBCT, ALL\_AML and MLL respectively to find their optimal gene subsets. In contrast, MGRFE merely used 10.8230, 9.0108 and 8.8739 minutes respectively in the same three datasets and thus showed much higher converge speed. Moreover, according to Tables 4, 5 and 6 in the main manuscript, the gene subsets selected by MGRFE had smaller sizes but higher classification accuracies compared with Kar \emph{et~al.}'s method. We noted that Kar \emph{et~al.} didn't employ the filter techniques to cut down the feature search space and their binary-coded PSO demanded high time cost to converge when dealing with the thousands of genes in each microarray.

@@

>>>>

	We implemented the proposed MGRFE in Python version 3.6.0 environment (\url{https://www.python.org/}) on a common laptop computer with Intel(R) Core(TM) i5-4210U CPU and 8G memory. The Python SciPy package version 0.19.0 \cite{RN441} was involved in the \textit{t}-test process, and minepy package version 1.2.0 \cite{RN440} was used to perform the MIC calculation. Some parameter settings about MGRFE: 1) the evolution iteration number of embedded GA was dynamically set to 1 to 3 (smaller iteration number used for larger chromosome length to save time); 2) the reduced feature number between two GA runs, the RFE step, was dynamically set to 1 to 3 (larger reduction step used for larger chromosome length to save time); and 3) the iterative layer number of MGRFE being 3 with three, two and one GaRFE processes at each layer is usually enough. In the experiments, we limited the size of the final selected gene combination in each dataset to below 10 genes. According to the experiment records, in each of the 19 microarray datasets, the running time of MGRFE is commonly between 500 seconds (8.33 minutes) and 900 seconds (15 minutes). The running time has included the whole filter screen and later wraper search processes. Additionally, it is well worthy to mention that the final chosen gene subset in each dataset might be already found by the first GaRFE process in the first layer of MGRFE, which just costs 2\(\scriptsize{\sim}\)3 minutes. More implementation details and experiment results of MGRFE in the 19 datasets are available at \url{https://github.com/Pengeace/MGRFE-GaRFE}.

	Because Kar \emph{et~al.} also employed an evolutionary-computation method PSO, which is similar to GA, to select minimal informative genes in microarray and provided their program running time records on three datasets SRBCT, ALL\_AML, and MLL \cite{W15}, here, we offer a simple running time comparison between their method and MGRFE. Their PSO-based method cost 2.7956, 2.7906 and 7.1488 hours on SRBCT, ALL\_AML and MLL respectively to find their optimal gene subsets. In contrast, MGRFE merely used 10.8230, 9.0108 and 8.8739 minutes respectively in the same three datasets and thus showed much higher converge speed. Moreover, according to Tables 4, 5 and 6 in the main manuscript, the gene subsets selected by MGRFE had smaller sizes but higher classification accuracies compared with Kar \emph{et~al.}'s method. We noted that Kar \emph{et~al.} didn't employ the filter techniques to cut down the feature search space and  thus their binary-coded PSO demanded high time cost to converge when dealing with the thousands of genes in each microarray.

@@

	\section*{S6. The statistically significant genes in \emph{t}-test}
	\addcontentsline{toc}{section}{S6. The statistically significant genes in \emph{t}-test}

	In \emph{t}-test, this study adopt the widely used $p = 0.05$ significance threshold to select the differentially expressed genes with \emph{p}-values lower than the threshold.
	The chosen of \emph{p}=0.05 has also been experimentally validated by the sample distribution condition of 17 binary classification datasets and our final experiment results on these datasets. Table \ref{Tab: p-value} illustrates the number of significant genes with \emph{p}-value less than 0.05 in the \emph{t}-test. From Table \ref{Tab: p-value} we can note that, \emph{p}=0.05 is a relatively accommodative condition on 17 binary-class datasets, which can not only identify the most differentially expressed genes, but also avoid the inappropriately exclusion of too many genes.

	\begin{table*}[htbp]
	\centering
	\caption{Number of statistically significant features with \emph{t}-test-based \emph{p}-values less than 0.05 on 17 binary classification datasets.}
	\label{Tab: p-value}
		\begin{tabular}{l|lllllllll}
			\toprule
			Dataset & DLBCL & Pros & Colon & Leuk & Mye & ALL1 & ALL2 & ALL3 & ALL4\\
			\hline
			Significant features & 2632 & 5061 & 594 & 2449 & 1720 & 4387 & 644 & 571 & 1279\\
			Total features & 7129 & 12625 & 2000 & 7129 & 12625 & 12625 & 12625 & 12625 & 12625\\
			\midrule
			Dataset & CNS & Lym & Adeno & Gas & Gas1 & Gas2 & T1D & Stroke &\\
			\hline
			Significant features & 334 & 804 & 1799 & 8260 & 16454 & 15601 & 10159 & 5569 &\\
			Total features & 7129 & 4026 & 7457 & 22645 & 22283 & 22283 & 54675 & 54675 & \\
			\bottomrule
		\end{tabular}
	\end{table*}

	\section*{S7. Compare other filter methods with the \emph{t}-test and MIC combination}
	\addcontentsline{toc}{section}{S7. Compare other filter methods with the \emph{t}-test and MIC combination}

	In the feature space reduction stage, the study used \emph{t}-test and MIC for their efficiency and convenience in gene filtering process. The \emph{t}-test has been widely used and validated for detecting differentially expressed genes in microarray \cite{RN413, RN414, RN421}. But \emph{t}-test has limitation in dealing with multi-class dataset for multi-variate \emph{t}-test can't be performed directly. The recently proposed MIC shows excellent performance in detecting a wide range of associations in large datasets including microarray \cite{W8,W16}, and MIC can cope with multi-class dataset. Thus, we combined \emph{t}-test and MIC to complete the feature screen task.

	For the execution order, we perform the \emph{t}-test first and then MIC. By the \emph{p}=0.05 significance threshold in the \emph{t}-test, we can quickly find the statistically significant genes, which could notably reduce the gene feature range.
	Besides, it has been noticed that the MIC calculation is kind of time-consuming compared with \emph{t}-test, thus it is suitable to perform \emph{t}-test first to decrease the gene number.

	We also compared the performance of other filter method combinations with the \emph{t}-test+MIC combination, the result are shown in Table \ref{Table-com}.

	Firstly, the combination of first Anova then Fold change (FC), Anova+FC. The experiment was carried on 3 balanced datasets (Adeno, Gas1 and Pros) and 3 imbalanced datasets (DLBCL, Leuk, and CNS) using 5-flod cross validation. For Anova, the \emph{p}-value threshold was also set as 0.05 as in \emph{t}-test. From Table \ref{Table-com}, it can notice that with the combination of Anova+FC, the sizes of finally selected genes are 2, 4, and 8 on datasets Adeno, Leuk and CNS, respectivily. But by the \emph{t}-test+MIC combination, simply 1, 2 and 7 genes are needed to achieve the same performance on the 3 datasets. On the rest of 3 datasets, the two filter method combinations have similar performance. Thus, the filter method combination of Anova+FC is little inferior to the combination of \emph{t}-test+MIC in finding the minimal discriminative gene subset.

	Secondly, the combination of first “volcano plot” then MIC, Volcano plot+MIC. The “volcano plot” can combine the advantages of \emph{t}-test and fold-change, thus we use the “volcano plot” to replace the \emph{t}-test. The experiments were performed on 2 balanced datasets (Adeno and Pros) and 2 imbalanced datasets (DLBCL and Leuk) by 5-fold cross validation. According to experiment results in Table \ref{Table-com}, these two methods select same size of genes and achieve similar performance on all the tested 4 datasets.
	In the experiments, we noted one defect of “volcano plot” for gene selection in a range of different microarray datasets. When we use the “volcano plot” to selected informative genes, in each microarray dataset, we need to hand-tune the \emph{p}-value threshold in \emph{t}-test and fold-change threshold value in FC to obtain the satisfactory result. For example, on dataset Adeno, there are 894 informative genes have FC value larger or equal to 2; but on dataset Pros, the highest FC value in all genes is just 1.46. Thus, for different datasets, the threshold values in “volcano plot” should be different. In fact, for each of the tested 4 datasets, the threshold values in “volcano plot” have been hand-tuned individually and the finally assigned threshold values vary among different datasets. This situation pose difficulty for building automatically application on a wide range of microarray datasets. In contrast, the \emph{t}-test has the consistent \emph{p}-value setting in all the 17 datasets in experiments and shows more convenience.

	To conclude, the filter method combination of \emph{t}-test+MIC are more efficient and convenient than the Anova+FC or Volcano plot+MIC combinations for the feature range reduction task in our study.

	\begin{table*}[htbp]
		\centering\caption{The performance comparison among different filter method combinations by 5-fold cross validation}
		\resizebox{0.75\textwidth}{!}{
		\begin{threeparttable}
				\label{Table-com}
				\begin{tabular}{l|llllllllll}
					\toprule
					Filter Methods & Dataset & Genes & $Sn$ & $Sp$ & $Acc$ & $Avc$ & $MCC$ & $AUC$\\
					\midrule
					\multirow{6}{*}{\emph{t}-test+MIC}
					 & Adeno & 1 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0\\
					 & Gas1 & 3 & 0.986 & 0.973 & 0.980 & 0.980 & 0.961 & 0.99\\
					 & Pros & 4 & 0.980 & 0.982 & 0.981 & 0.981 & 0.963 & 0.98\\
					 & DLBCL & 3 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0\\
					 & Leuk & 2 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0\\
					 & CNS & 7 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0\\
					\hline
					\multirow{6}{*}{Anova+FC}
					 & Adeno & 2 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0\\
					 & Gas1 & 3 & 0.987 & 0.973 & 0.980 & 0.980 & 0.960 & 0.979\\
					 & Pros & 4 & 0.980 & 0.982 & 0.981 & 0.981 & 0.963 & 0.982\\
					 & DLBCL & 3 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0\\
					 & Leuk & 4 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0\\
					 & CNS & 8 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0\\
					\hline
					\multirow{4}{*}{Volcano plot+MIC}
					 & Adeno & 1 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0\\
					 & Pros & 4 & 0.980 & 0.982 & 0.980 & 0.981 & 0.963 & 0.968\\
					 & DLBCL & 3 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0\\
					 & Leuk & 2 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0\\
					\bottomrule
				\end{tabular}
				% \begin{tablenotes}

				% \end{tablenotes}
		\end{threeparttable}}
	\end{table*}

	\section*{S8. Independent validation of selected features}
	\addcontentsline{toc}{section}{S8. Independent validation of selected features}
	In this section, we validate the selected gene subsets from Leuk, Gas1 and Gas2 on independent datasets using 10-time 10-fold cross validation.

	The datasets Leuk and MLL both contain the sample data of leukemia subtypes ALL (acute lymphoblastic leukemia) and AML(acute myeloid). We use the dataset MLL to validate the selected genes in Leuk. On Leuk the selected gene probes are [\textit{M23197},\textit{M31523}], and the related genes are [\textit{CD33}, \textit{TCF3}]. For these two genes, the coresponding gene probes in MLL are [\textit{32874\_at}, \textit{36802\_at}, \textit{1373\_at}, \textit{1374\_g\_at}]. Thus, we test the classification performance of the obtained 4 gene probes on the ALL and AML samples in MLL dataset.

	The datasets Gas1 and Gas2 are both gastric cancer data but related to different gastric cancer subtypes. Gas1 is about non-cardia gastric cancer, while Gas2 is about cardia gastric cancer. These two datasets are both from ref. \cite{Y30} and have the same gene probes as features. The gene probes selected by MGRFE on Gas1 are [\textit{215380\_s\_at}, \textit{221928\_at},\textit{214746\_s\_at}], and the gene probes selected on Gas2 are [\textit{210125\_s\_at},\textit{206361\_at}]. For Gas1 and Gas2, we both validated the selected gene probe subset on the other dataset.

	From Table \ref{Tab:ind}, it can be noted that the selected gene features on Leuk achieved satisfying performance on MLL. The obtained accuracy is 0.963, just slightly lower than the classification accuracies achieved within the datasets MLL or Leuk. The gene subset selected in Gas1 and Gas2 also showed acceptable performance on the other dataset. The different gastric cancer subtypes could partially account for the performance decrease in these two datasets.

	\begin{table*}[htbp]
		\centering
		\caption{Independent validation of selected features from dataset Leuk, Gas1, and Gas2 by 10-time 10-fold CV.}
		\resizebox{0.75\textwidth}{!}{
			\begin{threeparttable}
				\label{Tab:ind}
				\begin{tabular}{ll|llllll}
					\toprule
					Feature test on & Feature from & Sn & Sp & Acc & Avc & MCC & AUC\\
					\midrule
					\textbf{MLL} & \textbf{Leuk} & 0.963 & 0.96 & 0.963 & 0.962 & 0.934 & 0.993\\
					MLL & MLL & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 \\
					Leuk & Leuk & 0.99 & 1.0 & 0.993 & 0.995 & 0.987 & 1.0 \\
					\hline
					Gas1 & Gas1 & 0.984 & 0.965 & 0.974 & 0.974 & 0.952 & 0.989\\
					\textbf{Gas1} & \textbf{Gas2} & 0.917 & 0.929 & 0.923 & 0.923 & 0.853 & 0.967\\
					\textbf{Gas2} & \textbf{Gas1} & 0.933 & 0.827 & 0.880 & 0.880 & 0.774 & 0.973\\
					Gas2 & Gas2 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0\\
					\bottomrule
				\end{tabular}
				\begin{tablenotes}\scriptsize
					\item For dataset MLL, only ALL and AML leukemia samples are taken into consideration in this experiment to stay consistent with Leuk dataset.
				\end{tablenotes}
		\end{threeparttable}}
	\end{table*}

	\clearpage
	\newpage
